Decisions. Decisions, Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package

### 2020-12-29 paper.rmd
clean up repo
fix typos

---
title             : "Decisions Decisions Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package"
shorttitle        : "LPA Best Practices"

author: 
  - name          : "Joshua Rosenberg"
    affiliation   : "1"
    corresponding : yes
    address       : "1122 Volunteer Blvd., Knoxville, TN, 37996"
    email         : "jmrosenberg@utk.edu"
  - name          : "Caspar van Lissa"
    affiliation   : "2"
  - name          : "Jennifer Schmidt"
    affiliation   : "3"
  - name          : "Patrick Beymer"
    affiliation   : "5"
  - name          : "Daniel Anderson"
    affiliation   : "4"
  - name          : "Matthew Schell"
    affiliation   : "3"

@@

abstract: 
keywords          : "Latent Profile Analysis, mixture models, finite mixture models, tutorial, R, MPlus, mclust"
wordcount         : "suggested to be no more than 3,000"

@@

In statistics classes, textbooks, and workshops, an example like the following
is common:

@@

In this tutorial, we aim to describe one of the most commonly-used---and
relevant to psychologists---application of the general mixture model, to cases
for which all of the variables for which (relatively) homogeneous groups are
identified from among a (relatively) heterogeneous sample are continuos, *latent
profile analysis* (LPA). After describing the method and some examples of its
use, we provide a tutorial for carrying out LPA in the context of a
freely-available, open-source statistical software package that we have
developed and supported over the past three years for R (R Core Team, 2019),
tidyLPA. Finally, we offer some ideas about best practices and informed
recommendations for researchers aiming to use LPA in their applied work, and
conclude with reflections on the role of statistical software--especially
software that is freely-available, open-source, and highly-performant--in the
psychological sciences.

@@

The goal of LPA is estimate the parameters for a number of distributions
(typically multivariate) from a single data set. Thus, such an approach is
model-based, and some descriptions in the literature refer to it as model-based
clustering (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
Raftery, 2017). Thus, one distinction between LPA and other, similar cluster
analytic approaches is that LPA is model-based; instead of using algorithms to
group together cases, LPA seeks to estimate parameters---in terms of variances
and covariances and how they are the same or different across profiles---that
best characterize the different distributions. Then, this approach seeks to
estimate a probability that the observation is a sample from the population
associated with each profile (or mixture component) for each observation in the
dataset.

@@

This package focuses on models that are commonly specified as part of LPA.
Because MPlus is so widely-used, it can be helpful to compare output from other
software to MPlus. The functions in tidyLPA that use mclust have been
benchmarked to MPlus for a series of simple models (with small datasets and for
models with small numbers of profiles. This R Markdown output contains
information on how mclust and Mplus compare. The R Markdown to generate the
output is also available here, and, as long as you have purchased MPlus (and
installed MplusAutomation), can be used to replicate all of the results for the
benchmark. Note that most of the output is identical, though there are some
differences in the hundreths decimal places for some. Because of differences in
settings for the EM algorithm and particularly for the start values (random
starts for MPlus and starting values from hierarchical clustering for mclust),
differences may be expected for more complex data and models.

One way that tidyLPA is designed to be easy to use is that it assumes a “tidy”
data structure (Wickham, 2014). This means that it emphasizes the use of a data
frame as both the primary input and output of functions for the package. Because
data is passed to and returned (in amended form, i.e., with the latent profile
probabilities and classes appended to the data) from the function, it makes it
easy to create plots or use results in subsequent analyses. Another noteworthy
feature of tidyLPA is that it provides the same functionality through two
different tools, one that is open-source and available through R, the mclust
package (Scrucca et al., 2017) and one that is available through the commercial
software MPlus (Muthen & Muthen, 1997-2017). Moreover, as both tools use the
same maximum likelihood estimation procedure, they are benchmarked to produce
the same output. Also, note that we have described the model specifications with
descriptions of what is estimated in terms of the variances and covariances, the
common names for the models (i.e., class-varying unrestricted), and the
covariance matrix associated with the parameterizationfor the six models that
are possible to be estimated.

@@

You can also install the development version of tidyLPA, which may have new
features that are not yet availalble on CRAN (because the version on CRAN is
periodically updated) from GitHub with:
``` {r gh-installation, eval = FALSE, echo = TRUE}
# install.packages("devtools")
devtools::install_github("data-edu/tidyLPA")
```
``` {r, message = F, echo = TRUE}
library(tidyLPA)
library(dplyr) # for preparing the data
```
## Plotting and Describing the Data

Here, we use a simulated dataset contains data on broad (general, and not situation- or topic-specific) interest, sense of enjoyment, and self-efficacy from the PISA assessment.
First, let's explore the data graphically using the {ggplot2} package:
```{r, echo = TRUE}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  tidyr::gather(key, val) %>% 
  ggplot(aes(x = val, group = key, fill = key)) + 
  geom_density() + 
  facet_wrap(~key) + 
  theme_bw()
```
We see that the data broadly _appears_ to be normally distributed, as the following statistics from the use of the psych package suggest:
``` {r, echo = TRUE}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  psych::describe()
```

@@

We can see a number of statistics for each of the models that were estimated,
including the AIC (Aikake information criterion; based on -2 \* the
log-likelihood, and penalized by number of parameters) and BIC (BIC: Bayesian
information criterion; based on -2 log-likelihood, and penalized by number of
parameters adjusted by sample size). Both are *penalized* likelihood statistics;
the log-likelihood will always decrease with the addition of additional
paremeters estimated by the model, whereas the AIC and BIC will account for the
number of parameters; lower values of either indicate better fit.

@@

In addition to the pragmatic challenges facing the analyst carrying out LPA, there 
are a number of statistical issues that anaylsts face. Past research is informative in
this respect; Bauer

@@

tp <- t %>% filter(use == 1) %>% mutate(`LPA Number` = 1:nrow(.)) %>%
  select( `LPA Number`, `GRoLTS Number` = Number, `Checklist Item`)

@@

Because LPA cannot accomodate any missing values (and so uses listwise deletion) for this recommendation, tidyLPA has a function for addressing missing data that are based upon single imputation (with two methods available, one using a contemporary, machine learning-based procedure).

@@

We provide a number of functions for model comparison. The `compare_solutions()` profile
takes as an argument a particular measure of fit (defaulting to the BIC), and returns information
for all of the models estimated. It also implements an ensemble, analytic hierarchiy approach, that uses weighted values of a number of fit indices; see [@akogul2017approach] for more.

@@

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

>>>>

---
title             : "Decisions. Decisions, Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package"
shorttitle        : "Using tidyLPA in accordance with best practices"

author: 
  - name          : "Joshua Rosenberg"
    affiliation   : "1"
    corresponding : yes
    address       : "1122 Volunteer Blvd., Knoxville, TN, 37996"
    email         : "jmrosenberg@utk.edu"
  - name          : "Caspar van Lissa"
    affiliation   : "2"
  - name          : "Patrick Beymer"
    affiliation   : "5"
  - name          : "Jennifer Schmidt"
    affiliation   : "3"
  - name          : "Daniel Anderson"
    affiliation   : "4"
  - name          : "Matthew Schell"
    affiliation   : "3"

@@

abstract: 
keywords          : "Latent Profile Analysis, mixture models, finite mixture models, tutorial, R, MPlus, mclust"
wordcount         : "5,385"

@@

In statistics classes and workshops, an example like the following
is commonly used:

@@

In this tutorial, we aim to describe one of the most commonly-used---and
relevant to psychologists---application of the general mixture model, to cases
for which all of the variables for which (relatively) homogeneous groups are
identified from among a (relatively) heterogeneous sample are continuous, *latent
profile analysis* (LPA). After describing the method and some examples of its
use, we provide a tutorial for carrying out LPA in the context of a
freely-available, open-source statistical software package that we have
developed and supported over the past three years for R (R Core Team, 2019),
tidyLPA. Finally, we offer some ideas about best practices and informed
recommendations for researchers aiming to use LPA in their applied work, and
conclude with reflections on the role of statistical software--especially
software that is freely-available, open-source, and highly-performant--in the
psychological sciences.

@@

The goal of LPA is estimate the parameters for a number of distributions
(typically multivariate) from a single data set. Thus, such an approach is
model-based, and some descriptions in the literature refer to it as model-based
clustering (Hennig, Meila, Murtagh, & Rocci, 2015; Scrucca, Fop, Murphy, &
Raftery, 2017). Thus, one distinction between LPA and other, similar cluster
analytic approaches is that LPA is model-based; instead of using algorithms to
group together cases, LPA seeks to estimate parameters---in terms of variances
and covariances and how they are the same or different across profiles---that
best characterize the different distributions. Then, this approach seeks to
estimate a probability that the observation is a sample from the population
associated with each profile (or mixture component) for each observation in the
data set.

@@

This package focuses on models that are commonly specified as part of LPA.
Because MPlus is so widely-used, it can be helpful to compare output from other
software to MPlus. The functions in tidyLPA that use mclust have been
benchmarked to MPlus for a series of simple models (with small data sets and for
models with small numbers of profiles. This R Markdown output contains
information on how mclust and Mplus compare. The R Markdown to generate the
output is also available here, and, as long as you have purchased MPlus (and
installed MplusAutomation), can be used to replicate all of the results for the
benchmark. Note that most of the output is identical, though there are some
differences in the hundredths decimal places for some. Because of differences in
settings for the EM algorithm and particularly for the start values (random
starts for MPlus and starting values from hierarchical clustering for mclust),
differences may be expected for more complex data and models.

One way that tidyLPA is designed to be easy to use is that it assumes a “tidy”
data structure (Wickham, 2014). This means that it emphasizes the use of a data
frame as both the primary input and output of functions for the package. Because
data is passed to and returned (in amended form, i.e., with the latent profile
probabilities and classes appended to the data) from the function, it makes it
easy to create plots or use results in subsequent analyses. Another noteworthy
feature of tidyLPA is that it provides the same functionality through two
different tools, one that is open-source and available through R, the mclust
package (Scrucca et al., 2017) and one that is available through the commercial
software MPlus (Muthen & Muthen, 1997-2017). Moreover, as both tools use the
same maximum likelihood estimation procedure, they are benchmarked to produce
the same output. Also, note that we have described the model specifications with
descriptions of what is estimated in terms of the variances and covariances, the
common names for the models (i.e., class-varying unrestricted), and the
covariance matrix associated with the parameterization for the six models that
are possible to be estimated.

@@

You can also install the development version of tidyLPA, which may have new
features that are not yet available on CRAN (because the version on CRAN is
periodically updated) from GitHub with:
``` {r gh-installation, eval = FALSE, echo = TRUE}
# install.packages("devtools")
devtools::install_github("data-edu/tidyLPA")
```
``` {r, message = F, echo = TRUE}
library(tidyLPA)
library(dplyr) # for preparing the data
```
## Plotting and Describing the Data

Here, we use a simulated data set contains data on broad (general, and not situation- or topic-specific) interest, sense of enjoyment, and self-efficacy from the PISA assessment.
First, let's explore the data graphically using the {ggplot2} package:
```{r, echo = TRUE}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  tidyr::gather(key, val) %>% 
  ggplot(aes(x = val, group = key, fill = key)) + 
  geom_density() + 
  facet_wrap(~key) + 
  theme_bw()
```
We see that the data broadly _appears_ to be normally distributed, as the following statistics from the use of the psych package suggest:
``` {r, echo = TRUE}
pisaUSA15 %>%
  select(broad_interest, enjoyment, self_efficacy) %>% 
  psych::describe()
```

@@

We can see a number of statistics for each of the models that were estimated,
including the AIC (Aikake information criterion; based on -2 \* the
log-likelihood, and penalized by number of parameters) and BIC (BIC: Bayesian
information criterion; based on -2 log-likelihood, and penalized by number of
parameters adjusted by sample size). Both are *penalized* likelihood statistics;
the log-likelihood will always decrease with the addition of additional
parameters estimated by the model, whereas the AIC and BIC will account for the
number of parameters; lower values of either indicate better fit.

@@

In addition to the pragmatic challenges facing the analyst carrying out LPA, there 
are a number of statistical issues that analysts face. Past research is informative in
this respect; Bauer

@@

tp <- t %>% filter(use == 1) %>% mutate(`LPA Number` = 1:nrow(.)) %>%
  select( `LPA Number`, `GRoLTs Number` = Number, `Checklist Item`)

@@

Because LPA cannot accommodate any missing values (and so uses listwise deletion) for this recommendation, tidyLPA has a function for addressing missing data that are based upon single imputation (with two methods available, one using a contemporary, machine learning-based procedure).

@@

We provide a number of functions for model comparison. The `compare_solutions()` profile
takes as an argument a particular measure of fit (defaulting to the BIC), and returns information
for all of the models estimated. It also implements an ensemble, analytic hierarchy approach, that uses weighted values of a number of fit indices; see [@akogul2017approach] for more.

@@

# References

\newpage

### 2020-12-29 paper.tex
fix typos

\documentclass[man]{apa6}

@@

            pdftitle={Decisions Decisions Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package},
            pdfauthor={Joshua Rosenberg, Caspar van Lissa, Jennifer Schmidt, Patrick Beymer, Daniel Anderson, \& Matthew Schell},

@@

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}

@@

\shorttitle{LPA Best Practices}
\affiliation{
\vspace{0.5cm}
\textsuperscript{1} University of Tennessee, Knoxville\\\textsuperscript{2} Utrecht University\\\textsuperscript{3} Michigan State University\\\textsuperscript{3} University of Oregon\\\textsuperscript{4} University of Wisconsin, Madison}
\keywords{Latent Profile Analysis, mixture models, finite mixture models, tutorial, R, MPlus, mclust\newline\indent Word count: suggested to be no more than 3,000}
\usepackage{csquotes}
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

\usepackage{longtable}
\usepackage{lscape}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage[flushleft]{threeparttable}
\usepackage{threeparttablex}

\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother

\title{Decisions Decisions Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package}
\author{Joshua Rosenberg\textsuperscript{1}, Caspar van Lissa\textsuperscript{2}, Jennifer Schmidt\textsuperscript{3}, Patrick Beymer\textsuperscript{5}, Daniel Anderson\textsuperscript{4}, \& Matthew Schell\textsuperscript{3}}
\date{}

\authornote{

Correspondence concerning this article should be addressed to Joshua Rosenberg, 1122 Volunteer Blvd., Knoxville, TN, 37996. E-mail: \href{mailto:jmrosenberg@utk.edu}{\nolinkurl{jmrosenberg@utk.edu}}}

\abstract{

@@

In this tutorial, we aim to describe one of the most commonly-used---and
relevant to psychologists---application of the general mixture model, to cases
for which all of the variables for which (relatively) homogeneous groups are
identified from among a (relatively) heterogeneous sample are continuos, \emph{latent
profile analysis} (LPA). After describing the method and some examples of its
use, we provide a tutorial for carrying out LPA in the context of a
freely-available, open-source statistical software package that we have
developed and supported over the past three years for R (R Core Team, 2019),
tidyLPA. Finally, we offer some ideas about best practices and informed
recommendations for researchers aiming to use LPA in their applied work, and
conclude with reflections on the role of statistical software--especially
software that is freely-available, open-source, and highly-performant--in the
psychological sciences.

@@

The goal of LPA is estimate the parameters for a number of distributions
(typically multivariate) from a single data set. Thus, such an approach is
model-based, and some descriptions in the literature refer to it as model-based
clustering (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \&
Raftery, 2017). Thus, one distinction between LPA and other, similar cluster
analytic approaches is that LPA is model-based; instead of using algorithms to
group together cases, LPA seeks to estimate parameters---in terms of variances
and covariances and how they are the same or different across profiles---that
best characterize the different distributions. Then, this approach seeks to
estimate a probability that the observation is a sample from the population
associated with each profile (or mixture component) for each observation in the
dataset.

@@

This package focuses on models that are commonly specified as part of LPA.
Because MPlus is so widely-used, it can be helpful to compare output from other
software to MPlus. The functions in tidyLPA that use mclust have been
benchmarked to MPlus for a series of simple models (with small datasets and for
models with small numbers of profiles. This R Markdown output contains
information on how mclust and Mplus compare. The R Markdown to generate the
output is also available here, and, as long as you have purchased MPlus (and
installed MplusAutomation), can be used to replicate all of the results for the
benchmark. Note that most of the output is identical, though there are some
differences in the hundreths decimal places for some. Because of differences in
settings for the EM algorithm and particularly for the start values (random
starts for MPlus and starting values from hierarchical clustering for mclust),
differences may be expected for more complex data and models.

One way that tidyLPA is designed to be easy to use is that it assumes a \enquote{tidy}
data structure (Wickham, 2014). This means that it emphasizes the use of a data
frame as both the primary input and output of functions for the package. Because
data is passed to and returned (in amended form, i.e., with the latent profile
probabilities and classes appended to the data) from the function, it makes it
easy to create plots or use results in subsequent analyses. Another noteworthy
feature of tidyLPA is that it provides the same functionality through two
different tools, one that is open-source and available through R, the mclust
package (Scrucca et al., 2017) and one that is available through the commercial
software MPlus (Muthen \& Muthen, 1997-2017). Moreover, as both tools use the
same maximum likelihood estimation procedure, they are benchmarked to produce
the same output. Also, note that we have described the model specifications with
descriptions of what is estimated in terms of the variances and covariances, the
common names for the models (i.e., class-varying unrestricted), and the
covariance matrix associated with the parameterizationfor the six models that
are possible to be estimated.

@@

You can also install the development version of tidyLPA, which may have new
features that are not yet availalble on CRAN (because the version on CRAN is
periodically updated) from GitHub with:

@@

Here, we use a simulated dataset contains data on broad (general, and not situation- or topic-specific) interest, sense of enjoyment, and self-efficacy from the PISA assessment.
First, let's explore the data graphically using the \{ggplot2\} package:

@@

We can see a number of statistics for each of the models that were estimated,
including the AIC (Aikake information criterion; based on -2 * the
log-likelihood, and penalized by number of parameters) and BIC (BIC: Bayesian
information criterion; based on -2 log-likelihood, and penalized by number of
parameters adjusted by sample size). Both are \emph{penalized} likelihood statistics;
the log-likelihood will always decrease with the addition of additional
paremeters estimated by the model, whereas the AIC and BIC will account for the
number of parameters; lower values of either indicate better fit.

@@

As missing data can have a bearing on the estimates, an important consideration is whether or not the missing data is associated with the levels of the variables---or not. Consider consulting general guidelines on missing data, including Enders (2010), Little \& Rubin (2019), and Allison (2001).
\#\#\# Is a description provided of what variables are related to attrition/missing data?

@@

We provide a number of functions for model comparison. The \texttt{compare\_solutions()} profile
takes as an argument a particular measure of fit (defaulting to the BIC), and returns information
for all of the models estimated. It also implements an ensemble, analytic hierarchiy approach, that uses weighted values of a number of fit indices; see (Akogul \& Erisoglu, 2017) for more.

@@

First, we note that developing a tool that emphasized both rigor and access has
been, so far, well-received. This suggests that there is an audience (or a
market) for tools to carry out sophisticated tools that may be otherwise
out-of-reach. Tools do not need to be hard to use, a point made by developers
of, for example, tools to carry out highly-sophisticated Bayesian methods using
the most cutting-edge tools (Bürkner \& others, 2017, Makowski, Ben-Shachar, Chen, and Lüdecke (2019)). We think that
tidyLPA speaks to this, too.

@@

@@

\leavevmode\hypertarget{ref-R-tidyLPA}{}%
Rosenberg, J. M., Beymer, P. N., Anderson, D. J., Lissa, C. van, \& Schmidt, J. A. (2018). TidyLPA: An r package to easily carry out latent profile analysis (lpa) using open-source or commercial software. \emph{Journal of Open Source Software}, \emph{3}(30), 978. doi:\href{https://doi.org/10.21105/joss.00978}{10.21105/joss.00978}

\leavevmode\hypertarget{ref-akogul2017approach}{}%
Akogul, S., \& Erisoglu, M. (2017). An approach for determining the number of clusters in a model-based cluster analysis. \emph{Entropy}, \emph{19}(9), 452.

\leavevmode\hypertarget{ref-R-tidyLPA}{}%
Rosenberg, J. M., Beymer, P. N., Anderson, D. J., Lissa, C. van, \& Schmidt, J. A. (2018). TidyLPA: An r package to easily carry out latent profile analysis (lpa) using open-source or commercial software. \emph{Journal of Open Source Software}, \emph{3}(30), 978. doi:\href{https://doi.org/10.21105/joss.00978}{10.21105/joss.00978}

>>>>

\documentclass[english,man]{apa6}

@@

            pdftitle={Decisions. Decisions, Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package},
            pdfauthor={Joshua Rosenberg1, Caspar van Lissa2, Patrick Beymer5, Jennifer Schmidt3, Daniel Anderson4, \& Matthew Schell3},

@@

@@

% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Using tidyLPA in accordance with best practices}
\keywords{Latent Profile Analysis, mixture models, finite mixture models, tutorial, R, MPlus, mclust\newline\indent Word count: 5,385}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{csquotes}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[shorthands=off,main=english]{babel}
\else
  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\fi

\title{Decisions. Decisions, Decisions: Carrying out Latent Profile Analysis in Accordance With Best Practices Using the tidyLPA R package}
\author{Joshua Rosenberg\textsuperscript{1}, Caspar van Lissa\textsuperscript{2}, Patrick Beymer\textsuperscript{5}, Jennifer Schmidt\textsuperscript{3}, Daniel Anderson\textsuperscript{4}, \& Matthew Schell\textsuperscript{3}}
\date{}

\affiliation{\vspace{0.5cm}\textsuperscript{1} University of Tennessee, Knoxville\\\textsuperscript{2} Utrecht University\\\textsuperscript{3} Michigan State University\\\textsuperscript{3} University of Oregon\\\textsuperscript{4} University of Wisconsin, Madison}

@@

In this tutorial, we aim to describe one of the most commonly-used---and
relevant to psychologists---application of the general mixture model, to cases
for which all of the variables for which (relatively) homogeneous groups are
identified from among a (relatively) heterogeneous sample are continuous, \emph{latent
profile analysis} (LPA). After describing the method and some examples of its
use, we provide a tutorial for carrying out LPA in the context of a
freely-available, open-source statistical software package that we have
developed and supported over the past three years for R (R Core Team, 2019),
tidyLPA. Finally, we offer some ideas about best practices and informed
recommendations for researchers aiming to use LPA in their applied work, and
conclude with reflections on the role of statistical software--especially
software that is freely-available, open-source, and highly-performant--in the
psychological sciences.

@@

The goal of LPA is estimate the parameters for a number of distributions
(typically multivariate) from a single data set. Thus, such an approach is
model-based, and some descriptions in the literature refer to it as model-based
clustering (Hennig, Meila, Murtagh, \& Rocci, 2015; Scrucca, Fop, Murphy, \&
Raftery, 2017). Thus, one distinction between LPA and other, similar cluster
analytic approaches is that LPA is model-based; instead of using algorithms to
group together cases, LPA seeks to estimate parameters---in terms of variances
and covariances and how they are the same or different across profiles---that
best characterize the different distributions. Then, this approach seeks to
estimate a probability that the observation is a sample from the population
associated with each profile (or mixture component) for each observation in the
data set.

@@

This package focuses on models that are commonly specified as part of LPA.
Because MPlus is so widely-used, it can be helpful to compare output from other
software to MPlus. The functions in tidyLPA that use mclust have been
benchmarked to MPlus for a series of simple models (with small data sets and for
models with small numbers of profiles. This R Markdown output contains
information on how mclust and Mplus compare. The R Markdown to generate the
output is also available here, and, as long as you have purchased MPlus (and
installed MplusAutomation), can be used to replicate all of the results for the
benchmark. Note that most of the output is identical, though there are some
differences in the hundredths decimal places for some. Because of differences in
settings for the EM algorithm and particularly for the start values (random
starts for MPlus and starting values from hierarchical clustering for mclust),
differences may be expected for more complex data and models.

One way that tidyLPA is designed to be easy to use is that it assumes a \enquote{tidy}
data structure (Wickham, 2014). This means that it emphasizes the use of a data
frame as both the primary input and output of functions for the package. Because
data is passed to and returned (in amended form, i.e., with the latent profile
probabilities and classes appended to the data) from the function, it makes it
easy to create plots or use results in subsequent analyses. Another noteworthy
feature of tidyLPA is that it provides the same functionality through two
different tools, one that is open-source and available through R, the mclust
package (Scrucca et al., 2017) and one that is available through the commercial
software MPlus (Muthen \& Muthen, 1997-2017). Moreover, as both tools use the
same maximum likelihood estimation procedure, they are benchmarked to produce
the same output. Also, note that we have described the model specifications with
descriptions of what is estimated in terms of the variances and covariances, the
common names for the models (i.e., class-varying unrestricted), and the
covariance matrix associated with the parameterization for the six models that
are possible to be estimated.

@@

You can also install the development version of tidyLPA, which may have new
features that are not yet available on CRAN (because the version on CRAN is
periodically updated) from GitHub with:

@@

Here, we use a simulated data set contains data on broad (general, and not situation- or topic-specific) interest, sense of enjoyment, and self-efficacy from the PISA assessment.
First, let's explore the data graphically using the \{ggplot2\} package:

@@

We can see a number of statistics for each of the models that were estimated,
including the AIC (Aikake information criterion; based on -2 * the
log-likelihood, and penalized by number of parameters) and BIC (BIC: Bayesian
information criterion; based on -2 log-likelihood, and penalized by number of
parameters adjusted by sample size). Both are \emph{penalized} likelihood statistics;
the log-likelihood will always decrease with the addition of additional
parameters estimated by the model, whereas the AIC and BIC will account for the
number of parameters; lower values of either indicate better fit.

@@

As missing data can have a bearing on the estimates, an important consideration is whether or not the missing data is associated with the levels of the variables---or not. Consider consulting general guidelines on missing data, including Enders (2010), Little and Rubin (2019), and Allison (2001).
\#\#\# Is a description provided of what variables are related to attrition/missing data?

@@

We provide a number of functions for model comparison. The \texttt{compare\_solutions()} profile
takes as an argument a particular measure of fit (defaulting to the BIC), and returns information
for all of the models estimated. It also implements an ensemble, analytic hierarchy approach, that uses weighted values of a number of fit indices; see (Akogul \& Erisoglu, 2017) for more.

@@

First, we note that developing a tool that emphasized both rigor and access has
been, so far, well-received. This suggests that there is an audience (or a
market) for tools to carry out sophisticated tools that may be otherwise
out-of-reach. Tools do not need to be hard to use, a point made by developers
of, for example, tools to carry out highly-sophisticated Bayesian methods using
the most cutting-edge tools (Bürkner \& others, 2017, @makowski2019). We think that
tidyLPA speaks to this, too.

@@

\newpage

@@

\leavevmode\hypertarget{ref-R-tidyLPA}{}%
Rosenberg, J. M., Beymer, P. N., Anderson, D. J., Lissa, C. van, \& Schmidt, J. A. (2018). TidyLPA: An r package to easily carry out latent profile analysis (lpa) using open-source or commercial software. \emph{Journal of Open Source Software}, \emph{3}(30), 978. \url{https://doi.org/10.21105/joss.00978}
